\chapter{Module A - Textual Conversations}

\section{Margaret Atwood} \label{09/02/2025}
	\textbf{What has Atwood identified as the influences, alterations, changes and inspirations for her work?}
	\begin{itemize}
		\item To answer the answered questions in the Tempest
		\begin{itemize}
			\item Epilogue "Set me free", to liberate from the prison of the play
			\item To find out what happened to Prospero
			\item Hag-seed is a derivation of \underline{The Tempest}, but is it also a text in its own right; a reimagining for a modern audience.
		\end{itemize}
	\item Linda Hutcheon asserts that an adaptation is "a derivation that is not derivative - a work that is second without being secondary. It is its own palimpsestic thing"
	\item Atwood states "adaptations require both a reverence for and willingness to desecrate their source material"
	\item This is the crux of the module: how different \textbf{contexts, values, and perspectives} account for the common and disparate aspects of \underline{Hag-seed}
	\end{itemize}

\newpage

\section{Context of The Tempest} \label{10/02/2025}
	\subsection{Context Summarising Activity}
		\begin{enumerate}
			\item Read the following handout and summarise your key understandings as a newspaper-style headline: a concise, memorable sentence or phrase. Aim to write 2 headlines for each section.
			\item For each headline you have written, find a quote from the play that you think is relevant to this understanding. Write it under your headline.
			\item Then write a sentence or two explaining how this quote from the play is connected to contextual matters. For example, you might think that the 'storm' in Act 1 Scene 1 can be read as a symbol of the dangers of disorder and the arguments on the sinking ship as an allegory for the dangers to society when nobody does their proper job, and everyone falls to chaos and squabbling.
		\end{enumerate}

	\subsection{The Tempest in its Context}
		\subsubsection{Why do we need to understand the historical and cultural context of the play in order to interpret it?}
			Of course we can read the play without knowing anything about the world that produced it. But the depth of our appreciation and understanding of the text is greatly increased if we can read 'between' and 'beyond' the lines and recognise the cultural assumptions and references that the story plays upon. Once we put the text into a context, we can begin to see it not merely as a lightweight and unconvincing fairy-tale but as a meaningful exploration of 'big ideas' and issues that were urgent in Shakespeare's time and are still relevant in our world today. Of course we can never know exactly what Shakespeare himself was trying to 'say' in this play (if he was trying to 'say' anything at all). But we can come close to understanding his intentions if we can reconstruct the attitudes and worldview that shaped the way the audiences of his own time would have interpreted his work.

			Their attitudes and worldview were very different from our own and it helps to be aware of this. For example, to our modern, liberal, democratic, feminist eyes, Prospero seems like a petty patriarchal dictator in his treatment of others. But pre-democratic England was not a liberal democracy. It was an authoritarian, hierarchical world of masters and servants. To people living in this harsher, stricter world, Prospero's stern and commanding character traits might not have seemed as extreme a flaw as they do to us today. Of course, the context of Shakespeare's own time isn't the only context that readers can use to bring meanings out of the play. Many contemporary critical readings and performances of The Tempest place its characters and events into modern contexts by reading it 'against the grain', as a critique of the patriarchy or of colonialism (see pp. 148 in your Cambridge School edition). But for our purposes in Module A it makes most sense to try to read the text as a product of its own time and place and a commentary on the world around it. This will allow us to bring out the contrast with Atwood's contemporary context and the newer attitudes and perspectives that she injects into her re-telling in Hag-Seed.

			\textbf{Activity}
				\begin{enumerate}
					\item Prospero angers the Left
					\item 
				\end{enumerate}

		\subsubsection{Historical Context}
			Shakespeare's first plays were written during the reign of Queen Elizabeth I (the 'Elizabethan' period in English history). She died, unmarried and childless, in 1603. Despite this unstable situation, to everyone's relief (and surprise) there was a peaceful handover of power. Elizabeth was succeeded by James I, a Scottish ruler from a new branch of the royal family called the Stuarts. Shakespeare's later plays, including The Tempest, therefore belong to the 'Jacobean' period (Jacobus being Latin for James).

			Shakespeare's Jacobean England - and indeed Europe as a whole - far from being a simple and stable society, was changing rapidly and profoundly. Old ways of thinking, inherited from the Middle Ages, were being challenged by new discoveries that were ushering in an entirely new understanding of the physical world, society and human nature. This era is called 'Early Modern' because it was the period in which the world we recognise today (a world of cities, trade, scientific knowledge, technological progress, capitalist economics, individualism and religious scepticism) first began - uncertainly and unevenly - to emerge.

			This was the 'Age of Discovery'. 'New Worlds' were being discovered, conquered and exploited by European voyagers, following in the wake of Columbus's discovery of the Americas in 1492. Europeans began to hear stories of alien peoples and cultures and to realise, for the first time, that their Christian Western way of life was not the only way of life possible. In 1584, Elizabeth's court was amazed when the explorer Walter Raleigh brought back two Native Americans, Wanchese and Manteo. This widening of cultural horizons fuelled fantasies of lands full of monsters, cannibals and fantastic beasts, greedy visions of El Dorado, and philosophical speculations about the possibilities of making a fresh start and building a utopian society somewhere beyond the seas. It also shook confidence in the account of Creation given in the scriptures.

			Discoveries were also being made at home. The advanced civilisation of pre-Christian Europe (Ancient Greece and Rome) was rediscovered by artists and intellectuals who began to call themselves 'humanists', because they believed that the proper study of mankind should be human nature and human society - not speculations about the afterlife. Humanist wisdom began to spread more widely and rapidly with the invention of the printing press and provided an alternative to traditional Church teachings. Humanism aimed to teach powerful people a curriculum of 'liberal arts' (foremost among which were grammar, rhetoric and logic, the ancestors of the subject English we do today) in the hope that it would make them into wiser and more virtuous rulers. Humanism's rediscoveries helped to spur on a new scientific spirit. Renaissance men like Galileo, Copernicus and Paracelsus had thrown off the limitations of religious orthodoxy and taken the first steps in the journey of modern scientific discovery. Dangerous knowledge was threatening the rule of the church. The very foundations of the universe were being shaken, as word spread that the earth was neither flat, nor the centre of the cosmos as mediaeval Christendom had long believed.

			Another disturbing change was also afoot. The first stirrings of modern capitalism had weakened the fixed social 'pyramid' of mediaeval feudalism. A new ambitious 'middle' class of 'self-made men', neither landowners nor peasants and increasingly rich, were beginning to challenge the power of the old aristocratic families and to discover the power of individual enterprise. These Early Modern businessmen built their fortunes, not on the mediaeval model (holding and inheriting farmland) but along modern lines, in the new international shipping trade and the banks that financed it.

			\textbf{Activity}
				\begin{enumerate}
					\item 
				\end{enumerate}

			Intellectual, cultural and economic upheavals were matched by political turmoil. In England, the new reigning monarch, James I, insisted that he ruled by 'divine right'. As God's Deputy on earth, he said, his decisions could not be questioned. He was, he said, like a Father or a Husband to the 'Island' over which he ruled - which meant he had complete patriarchal authority. But not everyone was convinced that the King's authority was absolute and unlimited. James was a terrible 'husband' to his state. He was largely indifferent to the needs of his people. Although he saw himself as a statesman, author and intellectual, he often avoided his royal duties to go hunting, drank heavily and was constantly in debt.

			His laws were continually blocked in parliament. High rents, stagnating wages and inflation in food prices in the countryside provoked regular peasant riots and a mass migration to London which led to a law-and-order panic as the city filled up with an underclass of beggars and vagrants. These were called 'masterless' men and their anarchic, ungoverned status caused much anxiety to respectable Englishmen. In 1605, Catholic terrorists plotted to blow up the King and Parliament with gunpowder. Most Jacobean observers feared that England was entering a phase of political decline and misrule.

			They were right. When James' son, Charles I, tried to command the same absolute obedience his father demanded, the country broke out into civil war. Charles was beheaded and the monarchy was - briefly - abolished in England. In this era of change, profound anxieties and doubts troubled the minds of artists and thinkers. To contemporary observers like the poet John Donne, it seemed like the whole world had gone 'out of joint'.

		\subsubsection{Performance Context}
			The Tempest was probably written in 1610-11 and most scholars think it was the last play that Shakespeare wrote before his retirement (not counting a couple of dodgy co-writing jobs). Some critics have seen the play's portrayal of the magician giving up the books and staff of his 'art' as a self-portrait by the playwright as he farewelled the theatre. We know this play was performed at the court of King James as part of a celebration for the engagement of his daughter to a German prince. The play's sound effects, extravagant costumes, musical accompaniment and masque scene might have been designed to appeal to the tastes of rich, jaded noble audiences. But the play was also performed for the middle class at the small indoor Blackfriars theatre and the common folk in Shakespeare's larger, cheaper outdoor theatre-in-the-round, the Globe (which seems to get a shout-out in Prospero's 'we are such stuff as dreams are made on' speech). We can assume that its portrayal of drunken servants mistaking themselves for kings, as well as the unflattering satirical depiction of corrupt courtiers like Antonio, might well have tickled these lower-class audiences.


		\subsubsection{Religious Ideas in the Play}
			Despite recent upheavals in belief, The Tempest was written in an age of faith. In Shakespeare's
			world, the Bible's teachings and sermons heard in church formed part of everyone's everyday mindset. The acts performed by the characters on his stage would have been judged by Christian standards of virtue and vice. Shakespeare's plays never deliver simple moral messages about sin and salvation, but they certainly explore moral issues within a broadly religious framework. The Tempest is full of sins, transgressions, boundary crossing and rule breaking - in the personal world and in the public and political arenas. Many of the characters commit crimes (particularly crimes of ambition and disobedience), which lead to a period of trial and torment, before finally being given the chance to repent of their sins and to seek God's forgiveness. For Christians, this sequence of sin, suffering, repentance and salvation is the underlying pattern of all human history and every human life. To Shakespeare's society, being a good and godly person meant discovering the following deep truths:
			\begin{itemize}
				\item Humans are flawed and sinful creatures by their nature;
				\item God sees, judges and punishes sin;
				\item But he also gives sinners a chance of redemption;
				\item If we repent, seek forgiveness and promise to sin no more we can attain heaven through Jesus' mercy.
			\end{itemize}
			But the play also makes reference to two other belief systems that uneasily cohabited in the minds
			of people in this time of transition: the new science and astronomy; and old superstitions and beliefs
			in magic and astrology. If the play seems confusing in its moral outlook, a jumble of different ways of
			thinking and acting, at times, this reflects the fact that it comes from a confused world where once
			unshakeable beliefs were no longer taken for granted.

		\subsubsection{Socio-Political Ideology in the Play}
			Every community is bound together by an 'ideology', a common system of 'ideas' (concepts an factual beliefs) and 'ideals' (values and visions) that everyone holds in common. The modern USA's ideology holds that wealth and power is won by merit and that anyone can go from humble origins to become President. The ideology of a selective school includes the ideas that some people are 'smart' by their nature and that intellectual 'giftedness' is accurately measured by pencil-and-paper tests. It includes the values of meritocracy and elitism. Obviously there is a gap between the ideological assumptions people use to interpret their world and the lived reality. For example, fervent believers in the American dream have to somehow overlook the reality that no woman has ever risen to the rank of President and that most of the country's rich inherited their wealth. The ideology of Elizabethan England, which was growing harder and harder to uphold in the face of change, emphasised the idea that all departments of God's creation was hierarchically ordered and functionally harmonious. From Heaven above to Hell below, within the cosmos, the social world, the natural environment and the human body, Being was organised into a ladder of higher and lower 'rungs' or a 'chain' with many links. Everything that existed had its rightful place or 'degree' in the 'Great Chain of Being' or the Scala Naturae. As Linley puts it: Most people knew where they were placed in the universal order, the Great Chain of Being. God ruled all, was omnipotent (all-powerful) and omniscient (all-knowing). Man was inferior to God, Christ, the Holy Ghost, all the angels, apostles, saints, the Virgin Mary and all the blessed, but superior to all animals, birds, fish, plants and minerals. God ruled Heaven, kings (and princes, dukes, counts, etc.) ruled on Earth and fathers ruled families, like God at home. The great hierarchical chain stretched from God through all the lesser hierarchies of existence to the very bottom in descending order of 
			importance - from divinity to dust - all interconnected as contributory parts of God's creation. The chain links were each a separate group of beings, creatures or objects, each connected to the one before and the one after, semi-separate, dependent but partly independent, separate yet part of something greater. Each link had its internal hierarchy. According to this ideology, the unequal social order, with the King in luxury at the top and the poor beggars in the street, was not an artificial construct, but a fact of nature, a God-given rule. The patriarchal family with the father as the Head of the family and the children as the Foot was likewise divinely ordained. Knowing your 'place', 'fearing God', serving your betters was a virtue. To sin was to overreach your place in the Great Chain, to disobey your betters or to slack off on your duties. Humans were placed below the angels because they were born sinful. Human beings too were a little hierarchy, with the virtues (our nobler selves) ruling over the 'animal appetites' (our baser instincts) below that tempted us to sin. People could hope to rise to the ranks of the 'blessed' in Heaven if they led a good life. Sins and vices were signs that humans were allowing themselves to fall down the ladder of existence towards the level of the beasts. It was thus understood that the Order was constantly under threat from Sin and Chaos. It therefore needed to be guarded carefully and constantly policed and reinforced. This was the ruler's obligation, the father's duty, the master's job: to be constantly vigilant, keep sin contained and the unruly lower orders in check. The powerful had to educate the less powerful. The higher-ups had to project and transmit a vision of the ideal order that the lower-downs could see and place their faith in. Anyone who was incapable of seeing this order, and refused to recognise their 'proper place' needed to be kept in check by constant surveillance and direct force if necessary. Shakespeare was always concerned with order and 'degree' - nationally, socially, personally and spiritually. Rebellion, usurpation and collapse are political themes found in many, perhaps all, of his plays. Disorder, excess and misrule figure in both his tragedies and comedies. Like many writers in the early 1600s, faced first with an unmarried heirless Queen and then with a feckless and distant James, he was preoccupied by political and social questions relating to how society should be run. Many of his characters give voice to a fear of the chaos that would ensue if the order broke down. Loss of 'degree' meant force would dominate society, justice would be lost, and illegitimate power, will and appetite ('an universal wolf') would rule. James I warned his son, 'Beware yee wrest not the World to your owne appetite, as over many doe, making it like A Bell to sound as yee please to interpret.'

			Shakespeare's most famous statement on the dangers of disorder and the importance of 'degree' is found in his play Troilus and Cressida, where the character Ulysses uses metaphors of untuned music, stormy skies and flooding seas (all very relevant in The Tempest) to describe what will happen if individuals allow give way to their own selfish, sinful desires and cease to observe their proper place in life. 

\newpage

\section{Revision of The Tempest} \label{12/02/2025}
	\subsection{Unjumble the Plot Activity}
		\begin{enumerate}
			\item Twelve years before the play begins, Prospero was the Duke of Milan.

			\item Prospero and Miranda land on an enchanted island inhabited by Caliban the son of the witch Sycorax, who had Ariel imprisoned in a pine tree until released by Prospero.

			\item Since then, Prospero has ruled over the island and its three occupants, Caliban, Ariel and Miranda, by the use of the magic arts. Caliban and Ariel are forced to obey Prospero, otherwise they are punished.

			\item As the play begins, Prospero discovers Antonio, Alonso, Sebastian and Alonso's son Ferdinand are in a boat nearby, and he uses his magic powers to raise a terrible storm that shipwrecks them on the island.

			\item Meanwhile, elsewhere on the island Miranda and Ferdinand meet and fall in love.

			\item Prospero makes Ariel torment the noblemen, separating Alonso and his son Ferdinand. Both believe the other to be dead.
			
			\item Sebastian plots with Antonio to murder Alonso and become King of Naples himself.
			
			\item Trinculo and Stephano - a jester and a butler - meet Caliban. They all get drunk and Caliban persuades them to join him to kill Prospero and thus rule the island.
			
			\item Prospero discovers Caliban's conspiracy with Trinculo and Stephano to kill him, and breaks it off.
			
			\item Having tested Ferdinand, Prospero accepts him as Miranda's future husband and presents the couple with a wedding masque.

			\item With the help of Ariel, Prospero gathers all his enemies together and forgives them.

			\item Prospero's brother Antonio, Alonso, the King of Naples, and his brother Sebastian exile Prospero, casting him and his baby daughter Miranda adrift on the roaring sea in a boat unfit for purpose.

			\item Prospero vows to abandon his magic arts.

			\item Prospero releases Ariel. Caliban and Prospero remain together on the island as everyone else leaves.
		\end{enumerate}

	\subsection{Genre Question}
		\textbf{Can The Tempest be understood as a tragedy of vengeance averted to resolve comically? If not, what genre does it fall into? Justify your response with at least two quotes from the play and two points from the genre readings you just completed.}

		Although \underline{The Tempest} demonstrates themes of vengeance, whereby Prospero seeks revenge on Alonso, the play ultimately does not fall into the category of a Shakespearean tragedy. As discussed in \textit{The Tempest as a Romance}, the play does not possess a credible plot. Compared to other tragedies like \underline{Macbeth}, the characters 

\section{Approach to Module A} \label{13/02/2025}
	\begin{itemize}
		\item Must have 50/50 discussion about \underline{The Tempest} and \underline{Hagseed}
		\item What ideas are generated in the first text
		\item What ideas carry over to the second text
		\item How does the reading of the second text impact our interpretation of the first text
	\end{itemize}

	\begin{enumerate}
		\item First Conversation: What does Text 1 say and what does Text 2 say similarly/differently
		\item Second Conversation: What does Text 1 say and what does Text 2 say similarly/differently
	\end{enumerate}

\section{Introduction to The Tempest}
	\subsection{Prior to reading}
		\begin{itemize}
			\item Last full play from Shakespeare, \underline{The Tempest} can be interpreted as a farewell to the theatre (Prospero's epilogue)
			\item Setting parallels the new world of America and contemporary London. Link to ideas of colonialism and challenge the new world with questions about Renaissance identity
			\begin{itemize}
				\item Free will
				\item Change in faith
				\item Explosion of arts
			\end{itemize}
			\item Modern special effects are able to bring more of the play to life - A spectacle
		\end{itemize}

	\subsection{Purpose/Intentions}
		Key verbs: invite, provoke, forewarn, unify, examine, re-examine, expose, criticise

	\subsection{Meaning/Main Ideas}
		\begin{itemize}
			\item What makes us Human?
			\begin{itemize}
				\item Persuading, influencing, advising, challenging, and ethical debate
			\end{itemize}
			\item Power of theatre
			\item Conflict between Humanism and Providentialism
			\item Political and patriarchal power
		\end{itemize}

	\subsection{Form of The Tempest}
			
